---
title: "Creating pedigree"
author: "david lopez idiaquez & devi satarkar"
date: "2026-04-23"
output: html_document
---
#0.Packages
```{r}
library(dplyr)
library(readxl)
library(readr)
```

#1. Loading datasets to create a list of the nestlings ringed in Wytham and their associated pnums (brood id -  to connect with adult data)

Note: we are only interested in those nestlings ringed in Wytham core area - Sections: mp, b, p, sw, c, o, ex, w, cp

Data on nestlings ringed in Wytham is a bit messy, so we are trying to maximise what we have from different datasets. Important to have the PNUMs to connect the nestling data to the dams and sires. 

##1.1 great_tit_nest_21_04_2026.csv

This dataset contains the nestlings ringed between 1960 and 2025 that David collated for the Life-History Evolution project with S.R. Evans. It was done by using two datasets long_term_ringing_data_latest_2026.02.19.csv and wythamGTBT_ringingdata_1960_2024.csv using the code recruits (all present in the repository),

This dataset should be a reliable source of information on the nestlings ringed between 1960 and 2025 and their associated Pnums
```{r}
great_tit_nest_21_04_2026 <- read_csv("great_tit_nest_21_04_2026.csv")
great_tit_nest_21_04_2026_tjoin<-great_tit_nest_21_04_2026 %>% 
  select(ring, Pnum, year)
```

##1.2 long_term_ringing_data_latest_2026.02.19.csv

From this one we can get information of the nestlings ringed pre 1960.
```{r}
long_term_ringing_data_latest_2026_02_19 <- read_csv("long_term_ringing_data_latest_2026.02.19.csv")

#filter the data
nest_pre1960<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(year<1960) %>%  #retain years pre 1960 
  filter(species=="greti") %>%  #retain great tits
  filter(age==1) #retain nestlings

#retain only those from relevant locations
nest_pre1960_2<-nest_pre1960 %>% 
  mutate(nest=substr(location, 6, 12)) %>%  #rationale: location=pnum and the first 5 characters are year + breeding attempt
  mutate(nest=trimws(nest),
         nest=toupper(nest))

table(nest_pre1960_2$nest) #there are some nestboxes named X something, which seem not to be within those in the core area
sum(is.na(nest_pre1960_2$nest)) #no nas in nest
sum(is.na(nest_pre1960_2$location)) #no nas in location


##load info on known box location to filter those from 
known_nestboxes_x_y <- read_csv("known_nestboxes_x_y.csv")

known_nestboxes_x_y_2<-known_nestboxes_x_y %>% 
  select(Nestbox, Section) %>% 
  mutate(Nestbox=trimws(Nestbox),
         Nestbox=toupper(Nestbox))

nest_pre1960_3<-nest_pre1960_2 %>% 
  left_join(known_nestboxes_x_y_2, by=c("nest"="Nestbox")) %>%
  mutate(Section = ifelse(is.na(Section), substr(nest, 1, 1), Section)) %>% 
  mutate(Section=trimws(Section),
         Section=tolower(Section)) %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp"))

nest_pre1960_4<-nest_pre1960_3 %>% 
  select(ring, location, year) %>% 
  rename(Pnum=location)
```

##1.3 join 1.1 & 1.2
```{r}
nestlings_from1947<-rbind(nest_pre1960_4, great_tit_nest_21_04_2026_tjoin)
```

#2. I now add the parents
##2.1 Adding dams and sires from Wytham GT BT 1960_2025.csv
This only contains information from 1960 onwards
```{r}
Wytham_GT_BT_1960_2025 <- read_csv("Wytham GT BT 1960_2025.csv")
adults_1960_dams<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Mother) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Dam=toupper(Mother),
         Dam=trimws(Dam)) %>% 
  na.omit() %>% 
  select(-Mother)

adults_1960_sires<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Father) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Sire=toupper(Father),
         Sire=trimws(Sire)) %>% 
  na.omit() %>% 
  select(-Father)
```

##2.2 Join mother and fathers from the breeding dataset
```{r}
nestlings_from1947_2<-left_join(nestlings_from1947, adults_1960_dams, by="Pnum")
nestlings_from1947_3<-left_join(nestlings_from1947_2, adults_1960_sires, by="Pnum")
```

##2.3 More parents can be Id'd from the long-term ringing data
###2.3.1 Sires
I am going to filter those with age>1 and the adult captures
```{r}
adults_lt<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(species=="greti") %>% 
  filter(age>1) %>% 
  select(ring, location, sex, year)

#here I select the males
adults_lt_males<-adults_lt %>% 
  filter(sex=="M") %>% 
  rename(sire_lt=ring) %>% 
  rename(Pnum=location) %>% 
  select(Pnum, sire_lt, year)
  
#I am getting several many-to-many conections between this dataset and the nestling one. This means that there are repeated Pnums in adults_lt_males (there are in the nestling data but that is to be expected).

#To deal with this, I am going to select the Pnums in the nestling data and join them to the adult dataset. This allows me to id the relevant rows, with relevant information for the nestlings, in the adult dataset and check the repeated Pnums in this dataset

nestling_pnums<-nestlings_from1947_3 %>% 
  distinct(Pnum) %>% 
  mutate(as_nest="yes")

adults_lt_males_2<-left_join(adults_lt_males,cosita, by="Pnum")

adults_lt_males_2_relevant_Pnums<-adults_lt_males_2 %>% 
  filter(as_nest=="yes")

#as there are some repeated pnums I need to check whether the ring, and thus the sire, associated to them is the same. If it is I can retain just one obs per pnum and join this to the nestling data.
adults_lt_males_2_relevant_Pnums2<-adults_lt_males_2_relevant_Pnums %>%
  group_by(Pnum) %>%
  summarise(n_distinct_sire = n_distinct(sire_lt)) %>%
  filter(n_distinct_sire > 1)
```

------There are 8 with two different rings in Pnum-------

1. 19731B12                
Two different rings: BP11278 & BP11272 -> checking the long term dataset BP11278 was captured also in SW12 the day after. I am going to assume that BP11272 is the good sire

2. 19901XN9 
Two different rings: F491499 & F491500 - Here I cannot know which is the sire and which the dam

3. 19921O57
According to the breeding dataset this is the good sire - F952677 [which could imply that this (VH83814) is the Dam identified as such at other breeding attempts]

4. 20151O52
Here both adults have been identified using the breeding data

5. 20161C154  
Here Z032801 was Id as a female once, while TP42633 has been consistently id as a male. I am going to assume that TP42633 is the sire

6. 20191SW8               
There are two sire TS44673 & TR43261. In that Pnum the dam was id as TS44673 according to the long term ringing data. So TR43261 is the sire

7. 20201O53              
Here I have the mother and the father from the breeding data

8. 20231EX76  
TW41773 Dam identified in by the breeding data. The two sires [TY68319, PF71305] are difficult to tear apart.

###2.3.2 Join sires
```{r}
adults_lt_males_2_relevant_Pnums_2<-adults_lt_males_2_relevant_Pnums %>% 
  filter(!(Pnum == "19731B12" & sire_lt == "BP11278")) %>%  #dealing with 1 retaining BP11272
  filter(!(Pnum == "19731B12")) %>% #dealing with 2; i remove both as I cannot know which is the sire
  filter(!(Pnum == "19921O57" & sire_lt == "VH83814")) %>% #dealing with 3 retaining F952677
  filter(!(Pnum == "20151O52")) %>% #dealing with 4, i remove it as both have been id with the breeding data
  filter(!(Pnum == "20161C154" & sire_lt == "Z032801")) %>% #dealing with 5, i retain TP42633
  filter(!(Pnum == "20191SW8" & sire_lt == "TS44673")) %>% #dealing with 6, i retain TR43261
  filter(!(Pnum == "20201O53")) %>% #Dealing with 7, Mother and father identified thanks to the breeding data
  filter(!(Pnum == "20231EX76")) %>%  #Dealing with 8, sires difficult to tear appart so I remove both
  select(Pnum, sire_lt, year) %>% 
  distinct(Pnum, .keep_all = TRUE)

nestlings_from1947_4<-left_join(nestlings_from1947_3, adults_lt_males_2_relevant_Pnums_2, by="Pnum")

#how many new ids does including the long-term data add
nestlings_from1947_4 %>%
  distinct(Pnum, .keep_all = TRUE) %>% 
  filter(is.na(Sire) & !is.na(sire_lt)) %>%
  nrow()
#this adds 417 new sires

#Sanity check - test that in those that I have both information of sire (from the breeding dataset) and of sire_lt (from the long term data)

nestlings_from1947_4_test <- nestlings_from1947_4 %>%
  mutate(Sire=trimws(Sire),
         Sire=toupper(Sire),
         sire_lt=trimws(sire_lt),
         sire_lt=toupper(sire_lt)) %>% 
  mutate(sire_match = case_when(
    is.na(Sire) | is.na(sire_lt) ~ NA_character_,
    Sire == sire_lt ~ "yes",
    TRUE ~ "no"
  ))
```

There are two problematic ones
1. 19781CP22
In this one there seems to be a typo between the sire in the breeding data and that in teh long term data [KC962*2*3 vs KC962*3*3].
Both are males but 223 appears in the long term ringing data as being captured in February and November while 233 was captured during the breeding season, thus I think 233 is the good one

2. 20181EX20
Here, the data from the breeding dataset shows the same individual [VZ29475] as sire and dam. According to ebmp VZ29475 is a female, and the long term breeding data gives me Y836584 as a male.

Fix these issues and combine sire columns
```{r}
nestlings_from1947_5 <- nestlings_from1947_4 %>%
  mutate(Sire = case_when(
    Pnum %in% c("19781CP22", "20181EX20") ~ NA_character_,
    TRUE ~ Sire
  ))%>%
  mutate(Sire = coalesce(Sire, sire_lt)) %>% 
  select(ring, Pnum, year.x, Dam, Sire) %>% 
  rename(year=year.x)

```

###2.3.3 Dams






