---
title: "Creating pedigree"
author: "david lopez idiaquez & devi satarkar"
date: "2026-04-23"
output: html_document
---
#0.Packages
```{r}
library(dplyr)
library(readxl)
library(readr)
```

#1. Loading datasets to create a list of the nestlings ringed in Wytham and their associated pnums (brood id -  to connect with adult data)

Note: we are only interested in those nestlings ringed in Wytham core area - Sections: mp, b, p, sw, c, o, ex, w, cp

Data on nestlings ringed in Wytham is a bit messy, so we are trying to maximise what we have from different datasets. Important to have the PNUMs to connect the nestling data to the dams and sires. 

##1.1 great_tit_nest_21_04_2026.csv

This dataset contains the nestlings ringed between 1960 and 2025 that David collated for the Life-History Evolution project with S.R. Evans. It was done by using two datasets long_term_ringing_data_latest_2026.02.19.csv and wythamGTBT_ringingdata_1960_2024.csv using the code recruits (all present in the repository),

This dataset should be a reliable source of information on the nestlings ringed between 1960 and 2025 and their associated Pnums
```{r}
great_tit_nest_21_04_2026 <- read_csv("great_tit_nest_21_04_2026.csv")
great_tit_nest_21_04_2026_tjoin<-great_tit_nest_21_04_2026 %>% 
  select(ring, Pnum, year)
```

##1.2 long_term_ringing_data_latest_2026.02.19.csv

From this one we can get information of the nestlings ringed pre 1960.
```{r}
long_term_ringing_data_latest_2026_02_19 <- read_csv("long_term_ringing_data_latest_2026.02.19.csv")

#filter the data
nest_pre1960<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(year<1960) %>%  #retain years pre 1960 
  filter(species=="greti") %>%  #retain great tits
  filter(age==1) #retain nestlings

#retain only those from relevant locations
nest_pre1960_2<-nest_pre1960 %>% 
  mutate(nest=substr(location, 6, 12)) %>%  #rationale: location=pnum and the first 5 characters are year + breeding attempt
  mutate(nest=trimws(nest),
         nest=toupper(nest))

table(nest_pre1960_2$nest) #there are some nestboxes named X something, which seem not to be within those in the core area
sum(is.na(nest_pre1960_2$nest)) #no nas in nest
sum(is.na(nest_pre1960_2$location)) #no nas in location


##load info on known box location to filter those from 
known_nestboxes_x_y <- read_csv("known_nestboxes_x_y.csv")

known_nestboxes_x_y_2<-known_nestboxes_x_y %>% 
  select(Nestbox, Section) %>% 
  mutate(Nestbox=trimws(Nestbox),
         Nestbox=toupper(Nestbox))

nest_pre1960_3<-nest_pre1960_2 %>% 
  left_join(known_nestboxes_x_y_2, by=c("nest"="Nestbox")) %>%
  mutate(Section = ifelse(is.na(Section), substr(nest, 1, 1), Section)) %>% 
  mutate(Section=trimws(Section),
         Section=tolower(Section)) %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp"))

nest_pre1960_4<-nest_pre1960_3 %>% 
  select(ring, location, year) %>% 
  rename(Pnum=location)
```

##1.3 join 1.1 & 1.2
```{r}
nestlings_from1947<-rbind(nest_pre1960_4, great_tit_nest_21_04_2026_tjoin)
```

#2. I now add the parents
##2.1 Adding dams and sires from Wytham GT BT 1960_2025.csv
This only contains information from 1960 onwards
```{r}
Wytham_GT_BT_1960_2025 <- read_csv("Wytham GT BT 1960_2025.csv")
adults_1960_dams<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Mother) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Dam=toupper(Mother),
         Dam=trimws(Dam)) %>% 
  na.omit() %>% 
  select(-Mother)

adults_1960_sires<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Father) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Sire=toupper(Father),
         Sire=trimws(Sire)) %>% 
  na.omit() %>% 
  select(-Father)
```

##2.2 Join mother and fathers from the breeding dataset
```{r}
nestlings_from1947_2<-left_join(nestlings_from1947, adults_1960_dams, by="Pnum")
nestlings_from1947_3<-left_join(nestlings_from1947_2, adults_1960_sires, by="Pnum")
```

##2.3 More parents can be Id'd from the long-term ringing data
###2.3.1 Sires
I am going to filter those with age>1 and the adult captures
```{r}
adults_lt<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(species=="greti") %>% 
  filter(age>1) %>% 
  select(ring, location, sex, year)

#here I select the males
adults_lt_males<-adults_lt %>% 
  filter(sex=="M") %>% 
  rename(sire_lt=ring) %>% 
  rename(Pnum=location) %>% 
  select(Pnum, sire_lt, year)
  
#I am getting several many-to-many conections between this dataset and the nestling one. This means that there are repeated Pnums in adults_lt_males (there are in the nestling data but that is to be expected).

#To deal with this, I am going to select the Pnums in the nestling data and join them to the adult dataset. This allows me to id the relevant rows, with relevant information for the nestlings, in the adult dataset and check the repeated Pnums in this dataset

nestling_pnums<-nestlings_from1947_3 %>% 
  distinct(Pnum) %>% 
  mutate(as_nest="yes")

adults_lt_males_2<-left_join(adults_lt_males,nestling_pnums, by="Pnum")

adults_lt_males_2_relevant_Pnums<-adults_lt_males_2 %>% 
  filter(as_nest=="yes")

#as there are some repeated pnums I need to check whether the ring, and thus the sire, associated to them is the same. If it is I can retain just one obs per pnum and join this to the nestling data.
adults_lt_males_2_relevant_Pnums2<-adults_lt_males_2_relevant_Pnums %>%
  group_by(Pnum) %>%
  summarise(n_distinct_sire = n_distinct(sire_lt)) %>%
  filter(n_distinct_sire > 1)
```

------There are 8 with two different rings in Pnum-------

1. 19731B12                
Two different rings: BP11278 & BP11272 -> checking the long term dataset BP11278 was captured also in SW12 the day after. I am going to assume that BP11272 is the good sire

2. 19901XN9 
Two different rings: F491499 & F491500 - Here I cannot know which is the sire and which the dam

3. 19921O57
According to the breeding dataset this is the good sire - F952677 [which could imply that this (VH83814) is the Dam identified as such at other breeding attempts]

4. 20151O52
Here both adults have been identified using the breeding data

5. 20161C154  
Here Z032801 was Id as a female once, while TP42633 has been consistently id as a male. I am going to assume that TP42633 is the sire

6. 20191SW8               
There are two sire TS44673 & TR43261. In that Pnum the dam was id as TS44673 according to the long term ringing data. So TR43261 is the sire

7. 20201O53              
Here I have the mother and the father from the breeding data

8. 20231EX76  
TW41773 Dam identified in by the breeding data. The two sires [TY68319, PF71305] are difficult to tear apart.

###2.3.2 Join sires
```{r}
adults_lt_males_2_relevant_Pnums_2<-adults_lt_males_2_relevant_Pnums %>% 
  filter(!(Pnum == "19731B12" & sire_lt == "BP11278")) %>%  #dealing with 1 retaining BP11272
  filter(!(Pnum == "19731B12")) %>% #dealing with 2; i remove both as I cannot know which is the sire
  filter(!(Pnum == "19921O57" & sire_lt == "VH83814")) %>% #dealing with 3 retaining F952677
  filter(!(Pnum == "20151O52")) %>% #dealing with 4, i remove it as both have been id with the breeding data
  filter(!(Pnum == "20161C154" & sire_lt == "Z032801")) %>% #dealing with 5, i retain TP42633
  filter(!(Pnum == "20191SW8" & sire_lt == "TS44673")) %>% #dealing with 6, i retain TR43261
  filter(!(Pnum == "20201O53")) %>% #Dealing with 7, Mother and father identified thanks to the breeding data
  filter(!(Pnum == "20231EX76")) %>%  #Dealing with 8, sires difficult to tear appart so I remove both
  select(Pnum, sire_lt, year) %>% 
  distinct(Pnum, .keep_all = TRUE)

nestlings_from1947_4<-left_join(nestlings_from1947_3, adults_lt_males_2_relevant_Pnums_2, by="Pnum")

#how many new ids does including the long-term data add
nestlings_from1947_4 %>%
  distinct(Pnum, .keep_all = TRUE) %>% 
  filter(is.na(Sire) & !is.na(sire_lt)) %>%
  nrow()
#this adds 417 new sires

#Sanity check - test that in those that I have both information of sire (from the breeding dataset) and of sire_lt (from the long term data)

nestlings_from1947_4_test <- nestlings_from1947_4 %>%
  mutate(Sire=trimws(Sire),
         Sire=toupper(Sire),
         sire_lt=trimws(sire_lt),
         sire_lt=toupper(sire_lt)) %>% 
  mutate(sire_match = case_when(
    is.na(Sire) | is.na(sire_lt) ~ NA_character_,
    Sire == sire_lt ~ "yes",
    TRUE ~ "no"
  ))
```

There are two problematic ones
1. 19781CP22
In this one there seems to be a typo between the sire in the breeding data and that in teh long term data [KC962*2*3 vs KC962*3*3].
Both are males but 223 appears in the long term ringing data as being captured in February and November while 233 was captured during the breeding season, thus I think 233 is the good one

2. 20181EX20
Here, the data from the breeding dataset shows the same individual [VZ29475] as sire and dam. According to ebmp VZ29475 is a female, and the long term breeding data gives me Y836584 as a male.

Fix these issues and combine sire columns
```{r}
nestlings_from1947_5 <- nestlings_from1947_4 %>%
  mutate(Sire = case_when(
    Pnum %in% c("19781CP22", "20181EX20") ~ NA_character_,
    TRUE ~ Sire
  ))%>%
  mutate(Sire = coalesce(Sire, sire_lt)) %>% 
  select(ring, Pnum, year.x, Dam, Sire) %>% 
  rename(year=year.x)

```

###2.3.3 Dams
```{r}
#here I select the females
adults_lt_females<-adults_lt %>% 
  filter(sex=="F") %>% 
  rename(dam_lt=ring) %>% 
  rename(Pnum=location) %>% 
  select(Pnum, dam_lt, year)

adults_lt_females_2<-left_join(adults_lt_females,nestling_pnums, by="Pnum")

adults_lt_females_2_relevant_Pnums<-adults_lt_females_2 %>% 
  filter(as_nest=="yes")

#as there are some repeated pnums I need to check whether the ring, and thus the sire, associated to them is the same. If it is I can retain just one obs per pnum and join this to the nestling data.
adults_lt_females_2_relevant_Pnums2<-adults_lt_females_2_relevant_Pnums %>%
  group_by(Pnum) %>%
  summarise(n_distinct_dam = n_distinct(dam_lt)) %>%
  filter(n_distinct_dam > 1)
```

Pnums with repeated values

1. 19931B63

Two rings as dams: F938343 and F833132 -> I cannot identify which may be the one

2. 20131B36

Two rings as dams: Y839845 and X235021 -> On ebmp Y839845 appears as true female and X235021 as suspected. X235021 was faceplated. Thus I am going to retain Y839845 as the true dam.

3. 20171EX91 : These were properly identified using the breeding data

4. 20171O2: 

Two rings as dams: TS45475 and TR42314. This TR42314 appears as diurnal scan, thus I am going to assume that TS45475 is the true dam.

5. 20191C4:

Two rings as dams: TR43098 and VZ31205. TR43098 Identified as a male by keith. VZ31205 is the female according to ebmp, thus i retain VZ31205. as dam.

6. 20201B207

Two rings as dams: TY70133 and TV44280, this last one appears twice. -> I cannot identify which is the true one

7. 20211CP35

Two rings as dams: TY69574 and PF72110. PF72110 id'd using the broodpatch TY69574 is a male in ebmp, so I am going to retain PF72110 as female

8. 20221SW16

Two rings as dams: VZ32782 and VZ33278. In ebmp VZ32782 is the dam of that brood

9. 20231B76      

Two rings as dams: TY70819 and TY69492. In ebmp TY69492 appears as the dam of that brood

10. 20231EX11: both Id'd thanks to the breeding data

11. 20231SW15

Two rings as dams: PF71392 and VZ32782. VZ32782 is the dam, and PF71392 the sire according to ebmp.

###2.3.4 Join dams
```{r}
adults_lt_females_2_relevant_Pnums_2<-adults_lt_females_2_relevant_Pnums %>% 
  filter(!(Pnum == "19931B63")) %>%  #dealing with 1 remove both 
  filter(!(Pnum == "20131B36" & dam_lt == "X235021")) %>%  #dealing with 2 retaining Y839845
  filter(!(Pnum == "20171EX91")) %>% #dealing with 3; properly idd in with the breeding data
  filter(!(Pnum == "20171O2" & dam_lt == "TR42314")) %>% #dealing with 4 retaining TS45475
  filter(!(Pnum == "20191C4" & dam_lt == "TR43098")) %>% #dealing with 5, retain VZ31205
  filter(!(Pnum == "20201B207")) %>% #dealing with 6, i retain TP42633
  filter(!(Pnum == "20211CP35" & dam_lt == "TY69574")) %>% #dealing with 7, i retain PF72110
  filter(!(Pnum == "20221SW16" & dam_lt == "VZ33278")) %>% #Dealing with 8, i retain VZ32782
  filter(!(Pnum == "20231B76" & dam_lt == "TY70819")) %>%  #Dealing with 9, i retain TY69492
  filter(!(Pnum == "20231EX11")) %>% #Dealing 10 both id thanks to the breeding data
  filter(!(Pnum == "20231SW15" & dam_lt == "PF71392")) %>% #Dealing with 11, i retain VZ32782
  select(Pnum, dam_lt, year) %>% 
  distinct(Pnum, .keep_all = TRUE)

nestlings_from1947_6<-left_join(nestlings_from1947_5, adults_lt_females_2_relevant_Pnums_2, by="Pnum")

#how many new ids does including the long-term data add
nestlings_from1947_6 %>%
  distinct(Pnum, .keep_all = TRUE) %>% 
  filter(is.na(Dam) & !is.na(dam_lt)) %>%
  nrow()
#this adds 484 new sires

#Sanity check - test that in those that I have both information of sire (from the breeding dataset) and of sire_lt (from the long term data)

nestlings_from1947_6_test <- nestlings_from1947_6 %>%
  mutate(Dam=trimws(Dam),
         Dam=toupper(Dam),
         dam_lt=trimws(dam_lt),
         dam_lt=toupper(dam_lt)) %>% 
  mutate(dam_match = case_when(
    is.na(Dam) | is.na(dam_lt) ~ NA_character_,
    Dam == dam_lt ~ "yes",
    TRUE ~ "no"
  ))
```

There are three problematic ones

-20051W3: It seems that there is a typo between the one I pasted from the breeding data and that of the long term data, but both appear only once so i think is ok to retain the one we have.

-20061EX38: Here there is a typo one starts with M and the other with N. They only appear once so it should be alright.

-20101W42: Typo between 0 and O the one that is as Dam is the good one as it appears like that at other captures

-20191O49: Here it also seems that there is a typo ending on 8 or 9. the one with 8 seems to be the good one

-20181O48: Here it also seems that there is a typo ending on 8 or 9. the one with 8 seems to be the good one (same typo as before!)

-20231W11: TW vs TY, TY is the one appearing on ebmp, i need to change it

Fix these issues and combine dam columns
```{r}
nestlings_from1947_7 <- nestlings_from1947_6 %>%
  mutate(Dam = case_when(
    Pnum %in% c("20191O49", "20181O48", "20231W11") ~ NA_character_,
    TRUE ~ Dam
  ))%>%
  mutate(Dam = coalesce(Dam, dam_lt)) %>% 
  select(ring, Pnum, year.x, Dam, Sire) %>% 
  rename(year=year.x)
```

###2.3.5 There are a couple extra Dam/Sires that I can add from these last two checks.

19921O57 - the dam can be VH83814 (see line 158)

20191C4 -  the sire can be TR43098 (see line 272)

```{r}
nestlings_from1947_8 <- nestlings_from1947_7 %>%
  mutate(Dam = ifelse(Pnum == "19921O57", "VH83814", Dam)) %>% 
  mutate(Sire = ifelse(Pnum == "20191C4", "TR43098", Sire))
```

####2.3.5.1 Check for the presence of Dams as sires and viceversa
Dams as sires
```{r}
#dams as sires
dam_as_sire <- intersect(
  unique(na.omit(nestlings_from1947_8$Dam)),
  unique(na.omit(nestlings_from1947_8$Sire))
)

ped_dams_as_sires<-nestlings_from1947_8 %>% 
  filter(Dam %in% dam_as_sire) %>% 
  distinct(Dam)
```
1 TY69574 -> this one needs to be removed from the sire column
2 TK26998 -> this appears as male in the long term ringing data - remove it from the dam column
3 E985078 -> this appears as male in the long term ringing data - remove it from the dam column
4 Y031429 -> this appears as a female in the long term ringing data - remove it from the sire column
5 VZ30433 -> not sure, remove it from the sire column (appeas as father and mother)
6 TX42961 -> not sure, remove it from the sire column (appeas as father and mother)
7 TV44507 -> keith sexed it as a female, remove it from the sire column
8 VZ30547 -> keith sexed it as a male, remove it from the dam column
9 PY93410 -> is a sire, remove it from dam. In 20241MP28 the sire is PY93410 and the dam is VZ33466 according to ebmp

```{r}
str(nestlings_from1947_8)

nestlings_from1947_8.5 <- nestlings_from1947_8 %>%
  mutate(Sire = na_if(Sire, "TY69574")) %>% 
  mutate(Dam = na_if(Dam, "TK26998")) %>% 
  mutate(Dam = na_if(Dam, "E985078")) %>% 
  mutate(Sire = na_if(Sire, "Y031429")) %>% 
  mutate(Sire = na_if(Sire, "VZ30433")) %>% 
  mutate(Sire = na_if(Sire, "TX42961")) %>% 
  mutate(Sire = na_if(Sire, "TV44507")) %>% 
  mutate(Dam = na_if(Dam, "VZ30547")) %>%
  mutate(Dam = na_if(Dam, "PY93410")) %>%
  mutate(
    Sire = if_else(Pnum == "20241MP28", "PY93410", Sire),
    Dam  = if_else(Pnum == "20241MP28", "VZ33466", Dam)
  )

dam_as_sire2 <- intersect(
  unique(na.omit(nestlings_from1947_8.5$Dam)),
  unique(na.omit(nestlings_from1947_8.5$Sire))
) #0! Fixed

```

Sires as Dams
```{r}
sire_as_dam <- intersect(
  unique(na.omit(nestlings_from1947_8.5$Dam)),
  unique(na.omit(nestlings_from1947_8.5$Sire))
) #as expected, 0!
```

#3. Devi's corrections
Devi went through the pedigree and found the following missing data
  -K578593 (sire) and K578592 (dam) are stored as dummy IDs in your pedigree, they are parents of offspring from 20002O117 (ring numbers VF64293 to 8)

  -J644931 (dam)  and VE32654 (sire), stored as dummy IDs in your pedigree, parents of offspring from 20082W27 (ring numbers TL36433 to 7)

  -Y035995 (dam) for 20121SW12A with offspring ring numbers TP27869 to 73

```{r}
nestlings_from1947_8.5.5 <- nestlings_from1947_8.5 %>%
  mutate(Dam = ifelse(Pnum == "20002O117" & is.na(Dam), "K578592", Dam)) %>% 
  mutate(Sire = ifelse(Pnum == "20002O117" & is.na(Sire), "K578593", Sire)) %>% 
  mutate(Dam = ifelse(Pnum == "20082W27" & is.na(Dam), "J644931", Dam)) %>% 
  mutate(Sire = ifelse(Pnum == "20082W27" & is.na(Sire), "VE32654", Sire)) %>% 
  mutate(Dam = ifelse(Pnum == "20121SW12A" & is.na(Dam), "Y035995", Dam))
```



#4. Dummy IDs and cutting ped to 1960 onwards
  I am going to cut the pedigree to have from 1960 onward as there are many many NAs in the adult information before that. I have struggled finding adult information from the early years.

  I also need to create dummy identities for those missing parents
```{r}
nestlings_from1947_9<-nestlings_from1947_8.5.5 %>% 
  filter(year>=1960) %>% 
  mutate(
    Dam  = if_else(is.na(Dam),  paste0(Pnum, "_dam"),  Dam),
    Sire = if_else(is.na(Sire), paste0(Pnum, "_sire"), Sire)
  ) %>% 
  select(ring, Dam, Sire) %>% 
  rename(id=ring)
```

#5. Now I need to add population founders on the top of the pedigree

##5.1 Add Dams and Sires that do not appear as Ids in the top of the pedigree
For this I use pedtricks
```{r}
library(pedtricks)

nestlings_from1947_10<-fix_ped(nestlings_from1947_9)
```

##5.2 Add information from other individuals that appear in the breeding dataset and in the mass dataset
###5.2.1 Breeding dataset
```{r}
breeding_mothers<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  select(Mother) %>% 
  mutate(Mother=toupper(Mother),
         Mother=trimws(Mother)) %>% 
  distinct(Mother)%>% 
  rename(ring=Mother)

breeding_fathers<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  select(Father) %>% 
  mutate(Father=toupper(Father),
         Father=trimws(Father)) %>% 
  distinct(Father) %>% 
  rename(ring=Father)

breeding_adults<-rbind(breeding_mothers, breeding_fathers)

#join the information of the ids already present in the pedigree

id_ped<-nestlings_from1947_10 %>% 
  select(id) %>% 
  mutate(present_ped="yes")

breeding_adults_2<-left_join(breeding_adults, id_ped, by=c("ring"="id"))

breeding_adults_not_inped<-breeding_adults_2 %>% 
  filter(is.na(present_ped)) %>% 
  rename(id=ring) %>% 
  select(-present_ped) %>% 
  mutate(dam=NA,
         sire=NA)

nestlings_from1947_11<-rbind(nestlings_from1947_10, breeding_adults_not_inped)
```

###5.2.2 Body mass dataset
```{r}
adult_mass_breeding25_alt <- read_csv("adult_mass_breeding25_alt.csv")

mass_adults_ids<-adult_mass_breeding25_alt %>% 
  select(ring) %>% 
  distinct(ring)

id_ped_2<-nestlings_from1947_11 %>% 
  select(id) %>% 
  mutate(present_ped="yes")

mass_adults_ids_2<-left_join(mass_adults_ids, id_ped_2, by=c("ring"="id"))
mass_adults_ids_2_not_inped<-mass_adults_ids_2 %>% 
  filter(is.na(present_ped)) %>% 
  rename(id=ring) %>% 
  select(-present_ped) %>% 
  mutate(dam=NA_character_,
         sire=NA_character_)
  
nestlings_from1947_12<-rbind(nestlings_from1947_11, mass_adults_ids_2_not_inped)

nestlings_from1947_13<-nestlings_from1947_12 %>% 
  mutate(id=toupper(id),
         id=trimws(id),
         dam=toupper(dam),
         dam=trimws(dam),
         sire=toupper(sire),
         sire=trimws(sire)) %>% 
  distinct(id, .keep_all = TRUE)

#final
nestlings_from1947_14<-fix_ped(nestlings_from1947_13) #there is one row with all NAs

#write.csv(nestlings_from1947_14, "ped_updated_1960_2025.csv")
```

Note: As an extra test to see if all the individuals that should be included at least as id's are included, I downloaded the data from J.P. Woodman's paper -Age Specificity in Territory Quality and Spatial Structure in a Wild Bird Population- (AmNat 2025). He probably used an older version of the data that we used to create this pedigree. All individuals that he used are included in the pedigree, which is reasuring. 



