---
title: "Creating pedigree"
author: "david lopez idiaquez & devi satarkar"
date: "2026-04-23"
output: html_document
---
#0.Packages
```{r}
library(dplyr)
library(readxl)
library(readr)
```

#1. Loading datasets to create a list of the nestlings ringed in Wytham and their associated pnums (brood id -  to connect with adult data)

Note: we are only interested in those nestlings ringed in Wytham core area - Sections: mp, b, p, sw, c, o, ex, w, cp

Data on nestlings ringed in Wytham is a bit messy, so we are trying to maximise what we have from different datasets. Important to have the PNUMs to connect the nestling data to the dams and sires. 

##1.1 great_tit_nest_21_04_2026.csv

This dataset contains the nestlings ringed between 1960 and 2025 that David collated for the Life-History Evolution project with S.R. Evans. It was done by using two datasets long_term_ringing_data_latest_2026.02.19.csv and wythamGTBT_ringingdata_1960_2024.csv using the code recruits (all present in the repository),

This dataset should be a reliable source of information on the nestlings ringed between 1960 and 2025 and their associated Pnums
```{r}
great_tit_nest_21_04_2026 <- read_csv("great_tit_nest_21_04_2026.csv")
great_tit_nest_21_04_2026_tjoin<-great_tit_nest_21_04_2026 %>% 
  select(ring, Pnum, year)
```

##1.2 long_term_ringing_data_latest_2026.02.19.csv

From this one we can get information of the nestlings ringed pre 1960.
```{r}
long_term_ringing_data_latest_2026_02_19 <- read_csv("long_term_ringing_data_latest_2026.02.19.csv")

#filter the data
nest_pre1960<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(year<1960) %>%  #retain years pre 1960 
  filter(species=="greti") %>%  #retain great tits
  filter(age==1) #retain nestlings

#retain only those from relevant locations
nest_pre1960_2<-nest_pre1960 %>% 
  mutate(nest=substr(location, 6, 12)) %>%  #rationale: location=pnum and the first 5 characters are year + breeding attempt
  mutate(nest=trimws(nest),
         nest=toupper(nest))

table(nest_pre1960_2$nest) #there are some nestboxes named X something, which seem not to be within those in the core area
sum(is.na(nest_pre1960_2$nest)) #no nas in nest
sum(is.na(nest_pre1960_2$location)) #no nas in location


##load info on known box location to filter those from 
known_nestboxes_x_y <- read_csv("known_nestboxes_x_y.csv")

known_nestboxes_x_y_2<-known_nestboxes_x_y %>% 
  select(Nestbox, Section) %>% 
  mutate(Nestbox=trimws(Nestbox),
         Nestbox=toupper(Nestbox))

nest_pre1960_3<-nest_pre1960_2 %>% 
  left_join(known_nestboxes_x_y_2, by=c("nest"="Nestbox")) %>%
  mutate(Section = ifelse(is.na(Section), substr(nest, 1, 1), Section)) %>% 
  mutate(Section=trimws(Section),
         Section=tolower(Section)) %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp"))

nest_pre1960_4<-nest_pre1960_3 %>% 
  select(ring, location, year) %>% 
  rename(Pnum=location)
```

##1.3 join 1.1 & 1.2
```{r}
nestlings_from1947<-rbind(nest_pre1960_4, great_tit_nest_21_04_2026_tjoin)
```

#2. I now add the parents
##2.1 Adding dams and sires from Wytham GT BT 1960_2025.csv
This only contains information from 1960 onwards
```{r}
Wytham_GT_BT_1960_2025 <- read_csv("Wytham GT BT 1960_2025.csv")
adults_1960_dams<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Mother) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Dam=toupper(Mother),
         Dam=trimws(Dam)) %>% 
  na.omit() %>% 
  select(-Mother)

adults_1960_sires<-Wytham_GT_BT_1960_2025 %>% 
  filter(Species=="g") %>% 
  filter(Section %in% c("mp", "b", "p", "sw", "c", "o", "ex", "w", "cp")) %>% 
  select(Pnum, Father) %>% 
  mutate(Pnum=toupper(Pnum),
         Pnum=trimws(Pnum),
         Sire=toupper(Father),
         Sire=trimws(Sire)) %>% 
  na.omit() %>% 
  select(-Father)
```

##2.2 Join mother and fathers from the breeding dataset
```{r}
nestlings_from1947_2<-left_join(nestlings_from1947, adults_1960_dams, by="Pnum")
nestlings_from1947_3<-left_join(nestlings_from1947_2, adults_1960_sires, by="Pnum")
```

##2.3 More parents can be Id'd from the long-term ringing data

I am going to filter those with age>1 and the adult captures
```{r}
adults_lt<-long_term_ringing_data_latest_2026_02_19 %>% 
  filter(species=="greti") %>% 
  filter(age>1) %>% 
  select(ring, location, sex, year)

#here I select the males
adults_lt_males<-adults_lt %>% 
  filter(sex=="M") %>% 
  rename(sire_lt=ring) %>% 
  rename(Pnum=location) %>% 
  select(Pnum, sire_lt, year)
  
#I am getting several many-to-many conections between this dataset and the nestling one. This means that there are repeated Pnums in adults_lt_males (there are in the nestling data but that is to be expected).

#To deal with this, I am going to select the Pnums in the nestling data and join them to the adult dataset. This allows me to id the relevant rows, with relevant information for the nestlings, in the adult dataset and check the repeated Pnums in this dataset

nestling_pnums<-nestlings_from1947_3 %>% 
  distinct(Pnum) %>% 
  mutate(as_nest="yes")

adults_lt_males_2<-left_join(adults_lt_males,cosita, by="Pnum")

adults_lt_males_2_relevant_Pnums<-adults_lt_males_2 %>% 
  filter(as_nest=="yes")

#as there are some repeated pnums I need to check whether the ring, and thus the sire, associated to them is the same. If it is I can retain just one obs per pnum and join this to the nestling data.
adults_lt_males_2_relevant_Pnums2<-adults_lt_males_2_relevant_Pnums %>%
  group_by(Pnum) %>%
  summarise(n_distinct_sire = n_distinct(sire_lt)) %>%
  filter(n_distinct_sire > 1)

# # There are 8 with two different rings in Pnum
#   Pnum      n_distinct_sire
#   <chr>               <int>
# 1 19731B12                2 
##-two different rings: BP11278 & BP11272 -> checking the long term dataset BP11278 was captured also in SW12 the day after. I am going to assume that BP11272 is the good male

# 2 19901XN9                2 - This needs some thought
# 3 19921O57                2 - F952677 is the good one according to the breeding dataset
# 4 20151O52                2 - here I have the father and mother according to the breeding data
# 5 20161C154               2 - 
##here this Z032801 was once classified as a female so i think that I am going to assume that TP42633 is the male. I think this makes sense as the female was not id´d in that nest

# 6 20191SW8                2 - TR43261 is the female according to ebmp and TS44673 the male
# 7 20201O53                2 - here I have the mother and the father from the breeding data
# 8 20231EX76               2







  
nestlings_from1947_test<-left_join(nestlings_from1947_3, adults_lt_males, by="Pnum")


cosita<-nestlings_from1947_3 %>% 
  distinct(Pnum) %>% 
  mutate(as_nest="yes")

cosa<-left_join(adults_lt_males,cosita, by="Pnum")
str(test_count_pnums)
```












